Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-09-19 ms.tex
More edits

\begin{abstract}
  We calculate the steady-state hydrodynamic profiles of hot gas in
  the nuclei of a sample of early-type quiescent galaxies.  We assume
  that mass is supplied to the circumnuclear medium (CNM) entirely by
  stellar winds according to the radial stellar profiles for each
  galaxy, while the gas is heated by a combination of shocked stellar
  winds and supernovae.  Our analytic and numerical results are used
  to estimate the accretion rates onto the central supermassive black
  holes (SMBH) as a function of the SMBH mass $M_{\bullet}$ and the
  heating efficiency, the latter of which is related schematically to
  the age of the stellar population, the frequency of supernovae, and
  the efficacy of accretion feedback processes.  For a fixed heating
  efficiency, we find that the SMBH accretion Eddington factor
  increases with the SMBH mass, $\dot{M}/\dot{M}_{\rm edd} \propto
  M_{\bullet}^{\alpha}$ with $\alpha \approx 0.5, 1$ for cups and core
  nuclei, respectively.
  % However, considering the potentially lower heating efficiency of low
  % mass SMBHs (due to a younger stellar population and the rareity of
  % supernovae within the stagnation radius) and their resulting higher
  % susceptibility to cooling instability, we find a flatter
  % relationship $\alpha \approx 0$. 
  Our results are used to check previous estimates of the SMBH
  accretion rate assuming Bondi accretion and extrapolating the X-ray
  measured density and temperature to the unresolved Bondi radius.  We
  also discuss the implications of our results for the diversity of
  CNM environments encountered by the jetted outflows from tidal
  disruption events, and for the use of an assumed
  $\dot{M}-M_{\bullet}$ relationship to constrain the occupation
  fraction of SMBHs in low mass galaxies.
\end{abstract}

@@

The source term in the continuity equation (equation~\ref{eq:drhodt}),
%Should equations be referenced as equation of equation
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 10^{10}$ years (see, e.g.,
\citealt{Ciotti+91}).  In Appendix \ref{app:eta} we estimate that
$\eta \approx 0.11(t_{\star}/t_{\rm h})^{-1.26}$ for a Salpeter IMF,
assuming that all stars are formed in a burst of star formation at a
common age of $t_{\star}$ (equation~\ref{eq:eta}).  We adopt
$\eta=1$ as fiducial, but note that changing $\eta$ is equivalent to
rescaling $\rho$ by the same factor.\footnote{This is {\it not} true
  if cooling is included, since the cooling rate is proportional to
  density {\it squared}.}

Source terms $\propto q$ also appear in the momentum and entropy
equations (equation~\ref{eq:dvdt} and \ref{eq:dsdt}) due to the fact
that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy upon being incorporated
into the mean flow (see also below).  
%No net momentum is in the winds, but mass is being added which adds a
%term in the velocity equation...

@@

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability. This scheme is
described in \citealt{Brandenburg:2003a}.  We solve the equations for
different choices of $v_{\rm w,0}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM
structure, the time-dependent equations are solved to avoid numerical
issues that arise near the critical sonic points, of which there are
up to two because of the supersonic inflow on small radial scales and
a possible supersonic outflow on large scales.

@@

Figure~\ref{fig:profiles} shows three representative profiles of the
density $\rho(r)$, temperature $T(r)$, and radial velocity $|v(r)|$.
As expected, the gas density increases towards the SMBH.  For some
profiles (e.g. NGC1172 and NGC4478) the density $\rho$ follows a power
law from the stagnation radius (squares in Fig.~\ref{fig:profiles})
inwards $\rho \propto r^{n}$ with $n \approx 1$. Note that this is
shallower than the $-3/2$ for Bondi accretion. However, the power law
behavior does not extend through all radii. NGC4365 which has a
prominent break in the gas density profile, near the stagnation
radius. This is also roughly the location of the break in the stellar
light profile ($\rb$). The temperature profile is relatively flat at
large radii, but increases as $\propto 1/r^{k}$ interior to the sphere
of influence, where $k$ is smaller than 1 (which would be expected for
virialized gas).  The inwardly directed velocity increases towards the
hole approaching the expected free-fall rate $v \propto v_{\rm
  ff}\propto r^{-1/2}$ interior to the sonic point, and then increases
outwards towards large radii.
% {\bf BDM: I made this up - in general we need
%   more explanation of what the plots are actually showing, with some
%   physical explanation.}
% AG -- things are not quite approaching Bondi on the inner grid. I
% don't have a good physical explanation yet...
% AG -- I am somewhat concerned about boundary effects in
% NGC3115. Perhaps the break in the gas density profile is due to
% boundary effects--I know that in this case the break is not
% associated with a break in the stellar density profile

@@

This is shown in the top panel Figure~\ref{fig:bh_xray}. The
relationship from \citealt{MillerGallo+:2014a} and their detection
upper limits are shown as the blue and black solid lines
respectively. The points correspond to the $\dot{M} c^2$ for our
solutions (scaled by $5\times 10^{-7}$).  We must choose a $\vwO$ for
each galaxy.  We select $\vwO$=200 km/s if $\rs<\rIa$ for this $\vwO$,
$\vwO$=500 km/s if $\rs>\rIa$ for this $\vwO$, or exclude it. If
$L_x \sim \Mdot^2$ the discrepancy becomes worse, as shown in the
bottom panel of Fig.~\ref{fig:bh_xray}, where we plot $2 \times
10^{-4} (\eddr) \Mdot c^2$ from our solutions with the relation from
\citealt{MillerGallo+:2014a}.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{bh_xray.eps}
\caption{\label{fig:bh_xray} Accretion power plotted as a function of
  stellar mass for our solutions (points) and the trend of nuclear
  x-ray luminosity with stellar mass (black line) and the detection
  upper limit (blue line) from Fig. 1 of
  \citealt{MillerGallo+:2014a}. In the top panel we take the accretion
  power to be $\epsilon_1 \Mdot c^2$, with $\epsilon_1=5\times
  10^{-7}$ chosen to make our accretion powers comparable to the
  observed x-ray luminosities. In the bottom panel, we take the
  accretion power to be quadratic in $\Mdot$--$\epsilon_2 \eddr \Mdot
  c^2$ with $\epsilon2$=$2\times10^{-4}$. For a given galaxy, we take
  either $\vwO=200$ km/s (green points) or $\vwO=500$ km/s (orange
  points). We choose the solution for which the relative positions of
  $\rs$ and $\rIa$ are consistent with the assumed level of heating}
  %(see text for details).}
\end{figure}

@@

\subsection{Conduction}
\label{sec:conduction}{

@@

  \section{Summary and conclusions}
  \label{sec:summary}
  We have calculated steady-state models for circumnuclear medium of a
  sample of early type galaxies, assuming that gas is supplied
  entirely by stellar wind mass loss and heated by shocked stellar
  winds and SNe Ia. We may use these profiles to calculate an $\Mdot$
  onto each of sample galaxy's SMBH, and get the scaling of $\Mdot$
  with $\Mbh$. We may then compare our results with observed trends of
  nuclear x-ray luminosity. Our main conclusions may be summarized as
  follows.

  \begin{enumerate}
  \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value of
    $\alpha$ depends on the slope of the stellar density profile
    ($\alpha\simeq0.5,1$ for cusp and core galaxies respectively). One
    possible solution that the wind mass loss rate (i.e. $\eta$)
    decreases with $\Mbh$, as would be expected if low mass SMBH
    preferentially hosted older stellar populations. Another possible
    explanation is that the effective heating rate of the CNM $v_{\rm w,0}$
    decreases with decreasing SMBH mass. This could again be due to an
    older stellar population, since $v_{\rm w,0} \propto \eta^{−1/2}$
    for heating sources such as SN Ia and millisecond pulsars.
  \item We compare the $\Mdot$ for each of our solution to the one
    calculated from the Bondi formula. We find that they generally to
    a factor of a few. In our model the accretion rate is determined
    from the gas properties at the stagnation radius, $\rs$, which
    agrees with the Bondi radius, $\rb$, within a factor of a
    few. Thus, use of the Bondi formula is a good approximation.
  \item We compare our solutions for density and temperature to those
    inferred from Chandra x-ray observations by \citealt{Allen+:2006a}
    and \citealt{RussellMcNamara+:2013a} for a handful of galaxies in
    which our sample overlap with theirs. In general, we cannot match
    the observed gas density slope on large scales. 
  \end{enumerate}

>>>>

\begin{abstract}
  We calculate the steady-state hydrodynamic profiles of hot gas in
  the nuclei of a sample of early-type quiescent galaxies.  We assume
  that mass is supplied to the circumnuclear medium (CNM) entirely by
  stellar winds according to the radial stellar profiles for each
  galaxy, while the gas is heated by a combination of shocked stellar
  winds and supernovae.  Our analytic and numerical results are used
  to estimate the accretion rates onto the central supermassive black
  holes (SMBH) as a function of the SMBH mass $M_{\bullet}$ and the
  heating efficiency, the latter of which is related schematically to
  the age of the stellar population, the frequency of supernovae, and
  the efficacy of accretion feedback processes.  For a fixed heating
  efficiency, we find that the SMBH accretion Eddington factor
  increases with the SMBH mass, $\dot{M}/\dot{M}_{\rm edd} \propto
  M_{\bullet}^{\alpha}$ with $\alpha \approx 0.5, 1$ for cusp and core
  nuclei, respectively.
  % However, considering the potentially lower heating efficiency of low
  % mass SMBHs (due to a younger stellar population and the rareity of
  % supernovae within the stagnation radius) and their resulting higher
  % susceptibility to cooling instability, we find a flatter
  % relationship $\alpha \approx 0$. 
  Our results are used to check previous estimates of the SMBH
  accretion rate assuming Bondi accretion and extrapolating the X-ray
  measured density and temperature to the unresolved Bondi radius.  We
  also discuss the implications of our results for the diversity of
  CNM environments encountered by the jetted outflows from tidal
  disruption events, and for the use of an assumed
  $\dot{M}-M_{\bullet}$ relationship to constrain the occupation
  fraction of SMBHs in low mass galaxies.
\end{abstract}

@@

The source term in the continuity equation (equation~\ref{eq:drhodt}),
%Should equations be referenced as equation of equation
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 10^{10}$ years (see, e.g.,
\citealt{Ciotti+91}).  In Appendix \ref{app:eta} we estimate that
$\eta \approx 0.11(t_{\star}/t_{\rm h})^{-1.26}$ for a Salpeter IMF,
assuming that all stars are formed in a burst of star formation at a
common age of $t_{\star}$ (equation~[\ref{eq:eta}]).  We adopt
$\eta=1$ as fiducial, but note that changing $\eta$ is equivalent to
rescaling $\rho$ by the same factor.\footnote{This is {\it not} true
  if cooling is included, since the cooling rate is proportional to
  density {\it squared}.}

Source terms $\propto q$ also appear in the momentum and entropy
equations (equation~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) due to the fact
that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy upon being incorporated
into the mean flow (see also below).  
%No net momentum is in the winds, but mass is being added which adds a
%term in the velocity equation...

@@

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability. This scheme is
described in \cite{Brandenburg:2003a}.  We solve the equations for
different choices of $v_{\rm w,0}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM
structure, the time-dependent equations are solved to avoid numerical
issues that arise near the critical sonic points, of which there are
up to two because of the supersonic inflow on small radial scales and
a possible supersonic outflow on large scales.

@@

Figure~\ref{fig:profiles} shows three representative profiles of the
density $\rho(r)$, temperature $T(r)$, and radial velocity $|v(r)|$.
As expected, the gas density increases towards the SMBH.  For some
profiles (e.g. NGC1172 and NGC4478) the density $\rho$ follows a power
law from the stagnation radius (squares in Fig.~\ref{fig:profiles})
inwards $\rho \propto r^{n}$ with $n \approx 1$. Note that this is
shallower than the $-3/2$ power law for Bondi accretion. However, the
power law behavior does not extend through all radii, and the gas
density profile may have breaks mirroring those in the stellar density
profile. For example, NGC4365 has a prominent break in the gas density
profile, near the stagnation radius. This coincides with the location
of the break in the stellar light profile ($\rb$). The temperature
profile is relatively flat at large radii, but increases as $\propto
1/r^{k}$ interior to the sphere of influence, where $k$ is smaller
than 1--the power expected for virialized gas.  The inwardly
directed velocity increases towards the hole approaching the expected
free-fall rate $v \propto v_{\rm ff}\propto r^{-1/2}$ interior to the
sonic point, and then increases outwards towards large radii.
% {\bf BDM: I made this up - in general we need
%   more explanation of what the plots are actually showing, with some
%   physical explanation.}
% AG -- things are not quite approaching Bondi on the inner grid. I
% don't have a good physical explanation yet...
% AG -- I am somewhat concerned about boundary effects in
% NGC3115. Perhaps the break in the gas density profile is due to
% boundary effects--I know that in this case the break is not
% associated with a break in the stellar density profile

@@

This is shown in the top panel Figure~\ref{fig:bh_xray}. The
relationship from \citealt{MillerGallo+:2014a} and their detection
upper limits are the black and blue solid lines
respectively. The points correspond to the $\dot{M} c^2$ for our
solutions (scaled by $5\times 10^{-7}$).  We must choose a $\vwO$ for
each galaxy.  We select $\vwO$=200 km/s if $\rs<\rIa$ for this $\vwO$,
$\vwO$=500 km/s if $\rs>\rIa$ for this $\vwO$, or exclude it. If
$L_x \sim \Mdot^2$ the discrepancy becomes worse, as shown in the
bottom panel of Fig.~\ref{fig:bh_xray}, where we plot $2 \times
10^{-4} (\eddr) \Mdot c^2$ from our solutions with the relation from
\citealt{MillerGallo+:2014a}.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{bh_xray.eps}
\caption{\label{fig:bh_xray} X-ray luminosity plotted as a function of
  stellar mass for our solutions (points), the trend of nuclear
  x-ray luminosity with stellar mass (black line) and the detection
  upper limit (blue line) from Fig. 1 of
  \citealt{MillerGallo+:2014a}. In the top panel we take the X-ray
  luminosity to be $\epsilon_1 \Mdot c^2$, with $\epsilon_1=5\times
  10^{-7}$ chosen to make our luminosities comparable to those
  observed. In the bottom panel, we take the
  accretion power to be quadratic in $\Mdot$. $L_x=\epsilon_2 \eddr \Mdot
  c^2$ with $\epsilon2$=$2\times10^{-4}$. For a given galaxy, we take
  either $\vwO=200$ km/s (green points) or $\vwO=500$ km/s (orange
  points). We choose the solution for which the relative positions of
  $\rs$ and $\rIa$ are consistent with the assumed level of heating}
\end{figure}

@@

@@

  \section{Summary}
  \label{sec:summary}
  We have calculated steady-state models for circumnuclear medium of a
  sample of early type galaxies, assuming that gas is supplied
  entirely by stellar wind mass loss and heated by shocked stellar
  winds and SNe Ia. We may use these profiles to calculate an $\Mdot$
  onto each of sample galaxy's SMBH, and get the scaling of $\Mdot$
  with $\Mbh$. We may then compare our results with observed trends of
  nuclear x-ray luminosity. Our main conclusions may be summarized as
  follows.

  \begin{enumerate}
  \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value of
    $\alpha$ depends on the slope of the stellar density profile
    ($\alpha\simeq0.5,1$ for cusp and core galaxies respectively). One
    possible solution that the wind mass loss rate (i.e. $\eta$)
    decreases with $\Mbh$, as would be expected if low mass SMBH
    preferentially hosted older stellar populations. Another possible
    explanation is that the effective heating rate of the CNM $v_{\rm w,0}$
    decreases with decreasing SMBH mass. This could again be due to an
    older stellar population, since $v_{\rm w,0} \propto \eta^{−1/2}$
    for heating sources such as SN Ia and millisecond pulsars.
  \item We compare the $\Mdot$ for each of our solution to the one
    calculated from the Bondi formula. We find that they generally to
    a factor of a few. In our model the accretion rate is determined
    from the gas properties at the stagnation radius, $\rs$, which
    agrees with the Bondi radius, $\rb$, within a factor of a
    few. Thus, use of the Bondi formula is a good approximation.
  \item We compare our solutions for density and temperature to those
    inferred from Chandra x-ray observations by \citealt{AllenDunn+:2006a}
    and \citealt{RussellMcNamara+:2013a} for a handful of galaxies in
    which our sample overlap with theirs. In general, we cannot match
    the observed gas density slope on large scales. 
  \end{enumerate}
